---
title: "NICAR 2026- Follow bluesky users automatically"
output: html_document
date: "2026-02-23"
author: "Lucia Walinchus"
---

```{r setup, include=FALSE}
library(lubridate)
library(tidyverse)
library(ggplot2)
library(bskyr)
library(atrrr)
library(gutenbergr)

```


### Follow NICAR2026 users 

So you're excited to go to the [2026 NICAR data journalism conference](https://schedules.ire.org/nicar-2026/#/) and you want to follow all the speakers online. This quick tutorial will show you how do this with R and bluesky.

Note: you could also do a starter pack but I did not feel like searching for all those names. Trust me, this is easier. 


#### First you will have to create an app.

Go to Settings -> Privacy and Security -> App passwords and then click the big blue button that says "Add App password."

In your _console_, type _Sys.setenv(BLUESKY_PASS=_"put your APP password here") 


### Next, pull in the schedule. 

This only works with the JSON file, not the csv
```{r}
NICAR2026 <- rio::import("https://ire-nicar-conference-schedules.s3.us-east-2.amazonaws.com/nicar-2026/nicar-2026-schedule.json")

```

Make a dataframe of all the speakers 

```{r}
speakers <- NICAR2026$speakers
```

Get their socials
```{r}
socials <- speakers$social

```

Make that into a data frame
```{r}
get_socials <- bind_rows(socials)

```

And filter for bluesky

```{r}

bluesky <- get_socials %>% 
filter(label=="Bluesky") %>% 
  mutate(handle=stringr::str_replace(url,"https://bsky.app/profile/" ,""))
```



Then you can run this simple function I wrote to follow them all! 

```{r}
handles <- bluesky$handle

follow_everyone <- function(handle_to_follow) {
bs_follow(subject = handle_to_follow,
user = "walinchus.bsky.social",  #switch this out with your own handle
pass = Sys.getenv("BLUESKY_PASS"),
auth = bs_auth("walinchus.bsky.social", Sys.getenv("BLUESKY_PASS"), save_auth = NULL),
clean = TRUE)
Sys.sleep(sample(1:10, 1)) #don't overwhelm things
}
purrr::map(handles, follow_everyone)
```

